Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types

### 2016-02-24 manuscript.Rnw
bloom et al 2016
Clean trailing white space
Epic variable substitution
Define macros for common math terms

\bibliographystyle{newphyt}
\renewcommand{\figurename}{Fig.}
\usepackage[labelfont=bf, labelsep=quad]{caption}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

@@

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion. 
\begin{itemize}
\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm).  
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($M_F/M_S$) varied strongly between PFTs. We found that $M_F/M_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size. 
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged and managed plantations), excluding glasshouse and common garden studies. Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.

@@

Stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant. In all analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless 1.3m tall. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) by fitting the following equation, using non-linear regression.

\begin{equation}
D_S = D_{Sbh} H^c / (H - H_{bh})^c
\end{equation}

where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:

@@

\begin{equation}
\label{eq:mfmspart}
\frac{M_F}{M_S} = \frac{M_F}{A_F} \cdot \frac{A_F}{A_S} \cdot \frac{A_S}{M_S}
\end{equation}

where $M_F/A_F$ is the leaf mass per area (LMA), $A_F/A_S$ the ratio of plant leaf area to basal stem area, and $A_S/M_S$ the ratio of stem basal area to total woody above-ground mass. Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $M_F/M_S$ and for all three component variables.

@@

As expected, both whole-plant leaf mass ($M_F$) and above-ground woody biomass ($M_S$) showed a strong depence on plant height (Fig. 2). For both variables the relationship was highly non-linear on a log-log scale, demonstrating departure from simple allometric relationships. It is also evident that $M_S$ increases much quicker with plant height than $M_F$. As a result, the ratio $M_F/M_S$ quickly decreased with plant height (Fig. 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater 30m).

We found large differences between the three major PFTs in terms of above-ground biomass distribution (Fig. 3a). These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater evergreen angiosperm \textgreater deciduous angiosperm. Next we calculated $M_F/M_S$ at a common plant height for each of the PFTs. These three size-corrected $M_F/M_S$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($M_F/A_F$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit woody above-ground biomass ($A_F/M_S$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig. 3c), because the differences in $M_F/A_S$ were exactly compensated by differences in LMA (see also Fig. S2 for $A_F/M_S$ as a function of plant height).

Next, we decomposed $M_F/M_S$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}). For each of the component variables and $M_F/M_S$, we estimated the variance explained by plant height, PFT and climate variables (MAT or aridity) (Table~\ref{table:varparttable}). We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model. The first model included only plant height, and already explained much variation (65 - 79\%) in $M_F/M_S$ and $M_S/A_S$ but not LMA and $A_F/A_S$ (4 - 11 \%). When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $M_F/M_S$ and $M_F/A_F$ but not $A_F/A_S$ or $A_S/M_S$, demonstrating that PFT differences entered largely in leaf mass-based variables. Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig. 3b,c.

From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs. Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig. 4). We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.

Likewise, the ratio of stem basal area to stem mass ($A_S/M_S$) apparently varied little between PFTs (Table~\ref{table:varparttable}). To illustrate this further, a plot of $M_S$ as a function of $A_S$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a). The only exception was for large gymnosperm trees, which consistently showed a lower $M_S$ at a given $A_S$ (Fig. 5b). We fitted a GAM of $M_S$ as a function of $A_S$ (added as a smooth term) with or without PFT as a covariate. Again PFT was formally significant (likelihood ratio test, p \textless 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.

Taken together, these results demonstrate that 1) the majority of PFT-related variation in $M_F/M_S$ results from variation in LMA (not $A_F/A_S$ or $A_S/M_S$), and 2) the majority of height-related variation in $M_F/M_S$ results from variation in $A_S/M_S$ (not LMA or $A_F/A_S$). It is however important to note that there is very large variation between individuals even after accounting for PFT (Fig. 6). To correct for plant height effect on $M_S/A_S$, we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that $M_S$ is proportional to H and $A_S$ ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density). The histograms summarize that PFT effects are very evident in LMA (Fig. 6a), not in $A_F/A_S$, and only to some extent for $M_S/A_S$, which is entirely driven by large gymnosperm trees. However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table~\ref{table:varparttable}). The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study. When we tested for MAT effects on $M_F/M_S$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S5) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $M_F/M_S$. When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S6).

@@

Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate. Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the leaf mass per area. We found that at a common plant height, the leaf / woody biomass ratio ($M_F/M_S$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs. We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.

As expected, $M_F/M_S$ steeply declined with plant height. It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both $M_F$ and $M_S$ showed very non-linear patterns with plant height (Fig. 2), or among biomass pools (Fig. S4), even on a logarithmic scale. This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.

We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass. \citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI. Although we found remarkably similar leaf area per unit above-ground woody biomass ($M_F/M_S$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI. To scale our results to stand- and landscape scales thus requires more information on differences in stand structure between PFTs.

Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}. Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass. Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands. It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.

After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table~\ref{table:varparttable}). Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown). This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($M_F$) with MAT, when corrected for above-ground woody biomass ($M_S$). We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $M_F/M_S$ increased consistently with MAT for most size classes (Fig. S5), but not for gymnosperms. However, we found that for our database, this could be explained by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig. 1), which, as a result of higher LMA (Fig. 2) also have a higher $M_F/M_S$. However, based on a re-analysis of data in Reich et al. 2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $M_F/M_S$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.

@@

Global vegetation models assume or predict differences in biomass distribution between PFTs. However, these differences are currently poorly constrained, due to limited available empirical data. Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.

@@

<<results='asis'>>=
names(table_varpart_gam)[1] <- ""
print(xtable(table_varpart_gam,
             label="table:varparttable",
             caption="Explained variance in four plant biomass variables by plant height (H), plant functional type (PFT) and climate. Shown are adjusted $R^2$ estimated with generalized additive models (GAM), with four sets of predictors. All variables except PFT were added as smooth terms to the GAMs, avoiding the need to assume linear partial responses to the predictor variables, or search for best-fitting transformations. Climate variables included were either MAT or aridity index (calculated as the of PET to MAP). P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $M_F/M_S$ (p = 0.057)."),sanitize.text.function=function(x){x},
      include.rownames=FALSE, caption.placement = "top")

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \fi
    \caption{Raw data for leaf biomass ($M_F$) and above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table \ref{table:counttable}. Lines are generalized additive model fits to visualize the highly non-linear relationships. Note the logarithmic scale.}
    \label{fig:figrawdata}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \fi
    \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area. (a) Leaf mass per unit woody above-ground biomass ($M_F/M_S$) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from fitted models in panel (a). Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
    \fi
    \caption{Total plant leaf area ($A_F$) is approximately proportional to basal stem area ($A_S$), and this relationship is similar between PFTs. Lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.}
    \label{fig:alas}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
    \fi
    \caption{Above-ground woody biomass ($M_S$) scales with basal stem area ($A_S$), similarly for the three PFTs. (a) $M_S$ as a function of $A_S$ for the entire dataset. Lines are generalized additive model fits. (b) A close view of a portion of the data shown in panel (a), where evergreen gymnosperms displayed significantly lower woody biomass at a given stem basal area.}
    \label{fig:stemmassindex}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
    \fi
    \caption{Histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf area per unit basal stem area ($A_F/A_S$) and the stem mass index ($M_S/(A_S*H)$) grouped by the three PFTs. Arrows indicate means by PFT. Colours as in Fig. \ref{fig:figlmf}.}
    \label{fig:fighist}
\end{figure}

>>>>

\bibliographystyle{newphyt}
\renewcommand{\figurename}{Fig.}
\usepackage[labelfont=bf, labelsep=quad]{caption}
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\newcommand{\mf}{M_{\rm F}}
\newcommand{\ms}{M_{\rm S}}
\newcommand{\as}{A_{\rm S}}
\newcommand{\af}{A_{\rm F}}
\newcommand{\asb}{A_{\rm Sbh}}
\newcommand{\mlms}{\mf/\ms}
\newcommand{\asms}{\as/\ms}
\newcommand{\afas}{\af/\as}
\newcommand{\mfaf}{\mf/\af}
\newcommand{\afms}{\af/\ms}
\newcommand{\mfas}{\mf/\as}
\newcommand{\msas}{\ms/\as}
\newcommand{\mshas}{\ms/(H \cdot \as)}

@@

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
\begin{itemize}
\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm).
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($\mlms$) varied strongly between PFTs. We found that $\mlms$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size.
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged, and managed plantations), excluding glasshouse and common garden studies. Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because insufficient data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.

@@

Stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant. In all analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless 1.3m tall. For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) by fitting the following equation, using non-linear regression.

\begin{equation}
D_{\rm S} = D_{\rm Sbh} H^c / (H - H_{\rm bh})^c
\end{equation}

where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:

@@

\begin{equation}
\label{eq:mfmspart}
\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms}
\end{equation}

where $\mfaf$ is the leaf mass per area (LMA), $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to total woody above-ground mass. Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $\mlms$ and for all three component variables.

@@

As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong depence on plant height (Fig. 2). For both variables the relationship was highly non-linear on a log-log scale, demonstrating departure from simple allometric relationships. It is also evident that $\ms$ increases much quicker with plant height than $\mf$. As a result, the ratio $\mlms$ quickly decreased with plant height (Fig. 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater 30m).

We found large differences between the three major PFTs in terms of above-ground biomass distribution (Fig. 3a). These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater evergreen angiosperm \textgreater deciduous angiosperm. Next we calculated $\mlms$ at a common plant height for each of the PFTs. These three size-corrected $\mlms$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig. 3c), because the differences in $\mfas$ were exactly compensated by differences in LMA (see also Fig. S2 for $\afms$ as a function of plant height).

Next, we decomposed $\mlms$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}). For each of the component variables and $\mlms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table~\ref{table:varparttable}). We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model. The first model included only plant height, and already explained much variation (65 - 79\%) in $\mlms$ and $\msas$ but not LMA and $\afas$ (4 - 11 \%). When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mlms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables. Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig. 3b,c.

From the variance partitioning (Table 1), it is evident that $\afas$ did not vary appreciably between PFTs. Indeed, when all PFTs were plotted together, similar $\af$ was attained at a given $\as$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig. 4). We fit a GAM of $\af$ as a function of $\as$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.

Likewise, the ratio of stem basal area to stem mass ($\asms$) apparently varied little between PFTs (Table~\ref{table:varparttable}). To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a). The only exception was for large gymnosperm trees, which consistently showed a lower $\ms$ at a given $\as$ (Fig. 5b). We fitted a GAM of $\ms$ as a function of $\as$ (added as a smooth term) with or without PFT as a covariate. Again PFT was formally significant (likelihood ratio test, p \textless 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.

Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mlms$ results from variation in LMA (not $\afas$ or $\asms$), and 2) the majority of height-related variation in $\mlms$ results from variation in $\asms$ (not LMA or $\afas$). It is however important to note that there is very large variation between individuals even after accounting for PFT (Fig. 6). To correct for plant height effect on $\msas$, we expressed the stem mass ratio as $\mshas$ because we expect from basic geometry that $\ms$ is proportional to H and $\as$ ($\ms = \phi \rho \as H$, where $\phi$ a stem shape parameter, $\rho$ wood density). The histograms summarize that PFT effects are very evident in LMA (Fig. 6a), not in $\afas$, and only to some extent for $\msas$, which is entirely driven by large gymnosperm trees. However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table~\ref{table:varparttable}). The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study. When we tested for MAT effects on $\mlms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S5) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $\mlms$. When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S6).

@@

Read \citep{bloom2016}; global patterns in inferred allocation and biomass pools (leaf / stem).

Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate. Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the leaf mass per area. We found that at a common plant height, the leaf / woody biomass ratio ($\mlms$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs. We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.

As expected, $\mlms$ steeply declined with plant height. It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both $\mf$ and $\ms$ showed very non-linear patterns with plant height (Fig. 2), or among biomass pools (Fig. S4), even on a logarithmic scale. This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.

We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass. \citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI. Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mlms$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI. To scale our results to stand- and landscape scales thus requires more information on differences in stand structure between PFTs.

Previous interpretations of differences in $\mlms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}. Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass. Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands. It is however difficult to separate cause from effect in the relationships between LL, LMA, and $\mlms$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.

After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table~\ref{table:varparttable}). Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown). This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($\mf$) with MAT, when corrected for above-ground woody biomass ($\ms$). We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $\mlms$ increased consistently with MAT for most size classes (Fig. S5), but not for gymnosperms. However, we found that for our database, this could be explained by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig. 1), which, as a result of higher LMA (Fig. 2) also have a higher $\mlms$. However, based on a re-analysis of data in Reich et al. 2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $\mlms$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.

@@

Global vegetation models assume or predict differences in biomass distribution between PFTs. However, these differences are currently poorly constrained, due to limited available empirical data. Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $\mlms$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.

@@

<<results='asis'>>=
names(table_varpart_gam)[1] <- ""
print(xtable(table_varpart_gam,
             label="table:varparttable",
             caption="Explained variance in four plant biomass variables by plant height (H), plant functional type (PFT) and climate. Shown are adjusted $R^2$ estimated with generalized additive models (GAM), with four sets of predictors. All variables except PFT were added as smooth terms to the GAMs, avoiding the need to assume linear partial responses to the predictor variables, or search for best-fitting transformations. Climate variables included were either MAT or aridity index (calculated as the of PET to MAP). P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $\\mlms$ (p = 0.057)."),sanitize.text.function=function(x){x},
      include.rownames=FALSE, caption.placement = "top")

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \fi
    \caption{Raw data for leaf biomass ($\mf$) and above-ground woody biomass ($\ms$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table \ref{table:counttable}. Lines are generalized additive model fits to visualize the highly non-linear relationships. Note the logarithmic scale.}
    \label{fig:figrawdata}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \fi
    \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area. (a) Leaf mass per unit woody above-ground biomass ($\mlms$) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) $\mlms$ and leaf area ratio ($\afms$) at the average plant height in the dataset, estimated from fitted models in panel (a). Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
    \fi
    \caption{Total plant leaf area ($\af$) is approximately proportional to basal stem area ($\as$), and this relationship is similar between PFTs. Lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.}
    \label{fig:alas}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
    \fi
    \caption{Above-ground woody biomass ($\ms$) scales with basal stem area ($\as$), similarly for the three PFTs. (a) $\ms$ as a function of $\as$ for the entire dataset. Lines are generalized additive model fits. (b) A close view of a portion of the data shown in panel (a), where evergreen gymnosperms displayed significantly lower woody biomass at a given stem basal area.}
    \label{fig:stemmassindex}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs
    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
    \fi
    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$), leaf area per unit basal stem area ($\afas$) and the stem mass index ($\mshas$) grouped by the three PFTs. Arrows indicate means by PFT. Colours as in Fig. \ref{fig:figlmf}.}
    \label{fig:fighist}
\end{figure}
